A Tutorial on Tailored Simulation-Based Power Analysis for Experimental Designs with Generalized Linear Mixed Models

### 2024-04-08 docs/manuscript.Rmd
work on new practice sections on how to specify population parameters

### Strategies on specifying the population parameters

@@

##### Insightful descriptive statistics

For most GLMMs, the interpretation of model parameters if quite complicated which makes it difficult to specify plausible values for a specific application.
However, the mathematical structure of GLMMs determines which patterns in data would be produced by the model, if a specific set of values for the population parameters would be specified.
The knowledge of how to simulate from a GLMM enables us to compute insightful descriptive statistics that can be compared to available domain knowledge much more easily than the opaque values of model parameters.
For example, domain experts might not be able to specify plausible values for the $\beta$ coefficients in a logistic regression model (which are measured on the log-odds scale).
However, they should be able to reason about the expected ratio of the binary dependent variable in different experimental condition, i.e. which descriptive statistics they expect to observe (on average), when they will analyse the actual dataset produced by the study under planning (if everything turns out as postulated by the investigated theory). 
The job of the analyst who is familiar with the mathematical structure of the GLMM is to produce the model implied value of the insightful descriptive statistic expected by the domain expert.
Although insightful descriptive statistics often depend on the model parameters in complicated ways, it is usually possible to manually adjust the population parameters (if only with the help of trial and error) until the model implied quantities produce the desired result.

##### Insightful model based quantities

@@

#### Iterative process with domain experts

@@

@@

@@

@@

>>>>

## Strategies on specifying the population parameters

@@

Ideally, one would rely on meta-analytic results or conclusive data from pilot studies. However, these are sometimes not readily available. All parameter values in our present case study have been determined based on results from related study designs from the literature. Additionally, we had repeated discussions with our affiliated domain experts in radiology to check whether our assumptions seems plausible.

We now outline the strategies we used to determine plausible parameter values:
Unfortunately, the model parameters in a binomial GLMM are hard to interpret in isolation because 1) the $\beta$ parameters are connected to the modeled probability via the non-linear inverse-logit link, and 2) we also have to consider the random effects.
The most simple interpretation works by imagining a subject with average ability ($u_{0s} = 0$) responding to an item (i.e., CT scan) with average difficulty ($u_{0i} = 0$).
Then the model implied probability that such a person solves such an item accurately is given by:

$$
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
$$
In fact, we would only need the full equation if the subject is an expert and correct advice is presented.
In all other experimental conditions, some terms drop from the equation because they are multiplied by $0$.
The other extreme case would be the probability that a non-expert with average ability solves an item with average difficulty when no advice is presented:
$$
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
$$
We can revert this perspective by choosing plausible probability values and deriving the parameter values implied by these probabilities (that a subject with average ability solves an item with average difficulty) for each experimental condition.

### Insightful descriptive statistics

For most GLMMs, the interpretation of model parameters are quite complicated which makes it difficult to specify plausible values for a specific application.
However, the mathematical structure of GLMMs determines which patterns in data would be produced by the model, if a specific set of values for the population parameters would be specified.
The knowledge of how to simulate from a GLMM enables us to compute insightful descriptive statistics that can be compared to available domain knowledge much more easily than the opaque values of model parameters.
For example, domain experts might not be able to specify plausible values for the $\beta$ coefficients in a logistic regression model (which are measured on the log-odds scale).
However, they should be able to reason about the expected ratio of the binary dependent variable in different experimental condition, i.e. which descriptive statistics they expect to observe (on average), when they will analyse the actual dataset produced by the study under planning (if everything turns out as postulated by the investigated theory). 
The job of the analyst who is familiar with the mathematical structure of the GLMM is to produce the model implied value of the insightful descriptive statistic expected by the domain expert.
Although insightful descriptive statistics often depend on the model parameters in complicated ways, it is usually possible to manually adjust the population parameters (if only with the help of trial and error) until the model implied quantities produce the desired result.

In the last section, we showed how we can derive the model implied probability for each experimental condition, that a subject with average ability solves an item with average difficulty.
Although these derivations are straightforward, it is important not to misinterpret their implications: In binomial GLMMs, the average probability to solve an item (averaged across persons of varying ability and items of varying difficulty) is **not** equal to the probability that a person with average ability solves an item with average difficulty.
The first perspective implies a so-called marginal interpretation, while the second one implies a conditional interpretation.
For example, we determined the $\beta$ parameters in a way that corresponds to a desired conditional probability of $0.95$, that an expert with average ability solves an item with average difficulty when presented with correct advice.

However, even if the model were true, we would not observe that 95% of experts responding to items presented with correct advice from a big sample of subjects drawn from their natural distribution of ability and items drawn from their natural distribution of difficulty.
How much this estimated probability differs from the model implied probability of an average expert and an average items depends on the standard deviation of the random subject and random item intercepts (the two quantities are only equal if both standard deviations would be zero).
We can use the model implied observed proportion of correct diagnoses in each experimental condition as an insightful descriptive statistics to determine plausible values for the random effect standard deviations.

We will simulate a large dataset (for which the observed values of the descriptive statistic will be close to their model implied true values) and simply compute the relative frequency of correct diagnoses for each experimental conditions.

```{r}
library(tidyverse)
set.seed(1)
dat <- simulate(n_subjects = 3000, n_items = 3000,
  sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  group_by(condition) %>%
  summarize(relative_frequency = sum(y_bin) / n())
```

We can use these descriptive statistics to judge whether our chosen values for the random effect standard deviations would produce data that align with out domain expertise.

However, in our example they are not informative enough to determine a final set of plausible parameter values.
For this reason, we will additionally look at insightful model based quantities.

### Insightful model based quantities

@@

In the last section, we simulated a large simulated dataset and computed a descriptive statistic, the relative frequencies of correct diagnoses, for each experimental condition.
We will now use the model implied probability of each simulated data entry (y_prob) to visualize the whole model based implied marginal distribution of correct diagnoses for each condition.

We can use plots like the one above as a useful tool to decide whether the specified standard deviations of the subject and item random intercepts ($\sigma_S$ and $\sigma_I$) are reasonable by comparing the ranges and overlap between conditions to domain knowledge.

### Iterative process with domain experts

@@

```{r, echo=FALSE}
probtable <- data.frame(
  col1 = 
    c("no advice, no expert", "no advice, expert", "false advice, no expert", 
      "false advice, expert", "correct advice, no expert", "correct advice, expert"),
  col2 =
    c("0.70", "0.90", "0.40", "0.85", "0.90", "0.95"),
  col3 = 
    c("$logit(0.70) = \\beta_0$",
      "$logit(0.90) = \\beta_0 + \\beta_e$", 
      "$logit(0.40) = \\beta_0 + \\beta_a$",
      "$logit(0.85) = \\beta_0 + \\beta_e + \\beta_{a} + \\beta_{ea}$",
      "$logit(0.90) = \\beta_0 + \\beta_a + \\beta_c$",
      "$logit(0.95) = \\beta_0 + \\beta_e + \\beta_a + \\beta_c + \\beta_{ea} + \\beta_{ec}$"))
apa_table(
  probtable
  , col.names = c("Experimental condition",
    "$P(Y=1|\\mathbf{X=x}, \\mathbf{U} = \\mathbf{0})$",
    "Implied equation")
  , caption = "Assumed probabilities that an average subject solves an average item in each experimental condition."
  , note = "Implied equations are derived based on the model equations and setting all random intercept terms to 0."
  , escape = FALSE
)
```

@@

```{r}
library(tidyverse)
library(ggdist)
dat <- simulate(n_subjects = 3000, n_items = 3000, sd_u0s = 0.5, sd_u0i = 0.5)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0i = 0.01)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```

@@

```{r}
dat <- simulate(n_subjects = 2000, n_items = 2000, sd_u0s = 3, sd_u0i = 3)
dat %>% 
  mutate(condition = fct_cross(
    factor(expert), factor(advice_present), factor(advice_correct))) %>%
  mutate(condition = fct_recode(condition,
    "no expert, no advice" = "0:0:0", "expert, no advice" = "1:0:0", 
    "no expert, wrong advice" = "0:1:0", "expert, wrong advice" = "1:1:0",
    "no expert, correct advice" = "0:1:1", "expert, correct advice" = "1:1:1")) %>% 
  ggplot(aes(x = y_prob, y = condition)) +
  stat_histinterval(point_interval = "mean_qi", slab_color = "gray45") +
  scale_x_continuous(breaks = seq(0, 1, 0.1), limits = c(0, 1))
```
